Model Search in Relational Domains

### 2016-10-21 domains.tex
compile
domain comments

%-------------------------------------------------------------------------------
\subsection{Abstract Domain}
%-------------------------------------------------------------------------------
The work of~\cite{sas12,dhk2013-popl} shows that CDCL based SAT solvers 
operates on the partial assignment domains. We present a generic instantiation 
of CDCL architecture that can be extended to any relational and non-relational
domains.  
% 
In our prototype implementation and for illustrative purposes in this
paper we use a product domain $\mathcal{B}^b\times\mathcal{R}$ where
$\mathcal{B}$ is the Boolean domain
$\langle\{\true,\false,\bot,\top\},\Rightarrow,\wedge,\vee\rangle$, 
$b$ is the number of Boolean variables in the program, 
and $\mathcal{R}$ is a \emph{relational} domain over the 
numerical (bitvector) variables in their respective
data types (currently, signed and unsigned integers are supported).
%
We instantiate $\mathcal{R}$ with the template polyhedra domain.

For notational convenience we will denote elements of
$\mathcal{B}^b$ by their concretisations to propositional formulae,
e.g.\ for the Boolean variable vector $(g_1,c)$ with abstract value
$(\false,\top)$ we will write $\neg g_1$.

%-------------------------------------------------------------------------------
\paragraph{Template Polyhedra Abstract Domain}
%
The template polyhedra domain~\cite{sriram} represents sets $X$ of
valuations of a vector of variables $\vec{x}$ such that
$\mat{C}\vec{x}\leq\vec{d}$ for a fixed coefficient matrix $\mat{C}$
and a constant vector $\vec{d}$. The domain of $\vec{d}$ is augmented
by a special element to denote the minimal element $\bot$ of the lattice.
%
The abstraction function is defined as $\alpha(X) = \min \{\vec{d}\mid
\mat{C}\vec{x}\leq\vec{d}, \vec{x}\in X\}$ where $\min$ is
applied component-wise. There are several optimisation-based
techniques for computing the domain
operations~\cite{sriram,GS07b,BJKS15}. In our implementation we use
the strategy iteration approach available in 2LS~\cite{BJKS15}.
%
The concretisation $\gamma(\vec{d})$ is the set $\{\vec{x}\mid
\mat{C}\vec{x}\leq\vec{d}\}$; $\gamma(\bot)=\emptyset$.
%
For notational convenience we will use conjunctions of linear
inequalities, e.g.\ $x_1\geq 0 \wedge x_1-z\leq 30$ to denote elements
of the domain. $\true$ is $\top$ and $\false$ is $\bot$.

% \paragraph{Interval Abstract Domain} 
% The {\em Interval} abstract domain is a lattice 
% $ItvDom = (ItvElm, \sle_{I}, \join_{I}, \meet_{I})$, where
% $ItvElm: (Var \rightarrow Itv) \cup \{\bot\}$, and $Itv$ is 
% the set of intervals of type $[l,u]$ over numeric data 
% type with $l \leq u$. The least element is $\bot$ and the 
% greatest element is $\top$ which maps all variables to their
% minimum ($min$) and maximum ($max$) values.  An interval 
% $\langle x, [min, v] \rangle$ is written as $x \leq v$.  The 
% partial order $\sle_{I}$ over elements in the set $Itv$ is 
% given by $I_1 \sle_{I} I_2$ if $I_2$ contains $I_1$.
% A join $(\join_{I})$ of two intervals $\langle x_1 \rightarrow [l_1, u_1], 
% x_1 \rightarrow [l_2, u_2] \rangle$ is an interval 
% $\langle x_1 \rightarrow [min(l_1, l_2), max(u_1, u_2)]$.
% A meet $(\meet_{I})$ of two intervals $\langle x_1 \rightarrow [l_1, u_1], 
% x_1 \rightarrow [l_2, u_2] \rangle$ is an interval 
% $\langle x_1 \rightarrow [max(l_1, l_2), min(u_1, u_2)] \rangle$.
% The galois connection between the concrete environment and intervals is 
% given as follows.
% \[\alpha(x) = \{[inf(x), sup(x)] | x \in Var\} \qquad   \alpha(\emptyset) = \bot \]
% \[\gamma[p,q] = \{c \in \mathbb{Z} | p \leq c \leq q\} \qquad \gamma(\bot) = \emptyset \]
Template polyhedra subsume, for example, interval and octagon domains.

@@

The interval domain is a non-relational domain because a single
inequality only contains a single variable.
%
The octagon domain, however, is relational.  A fundamental difference
between relational and non-relational domains is that relational
domains require the computation of a \emph{closure} in order to obtain
a normal form that is usually required for efficient domain operations.
An example of a closure computation is shown below.
\[\mathit{closure}((x-y \leq 4) \wedge (y-z \leq 5))=((x-y \leq 4) \wedge (y-z \leq 5) \wedge (x-z \leq 9)) \]  
The closure computes all implied domain constraints.  For octagons,
the closure operation is very expensive (cubic complexity in the
number of program variables).
%
We will see that our algorithm performs closure computations lazily,
postponing them to the point where they are actually necessary.

@@

%-------------------------------------------------------------------------------
\paragraph{Abstract Transformers}
%
%In a typical abstract interpretation based Galois-connection setting
%over an over-approximate domain, every concrete element has a unique
%over-approximate representation in the abstract.  Likewise, every
%concrete transformer is over-approximated by a unique abstract
%transformer.  We now define an abstract deduction transformer.
%
An abstract transformer transforms an abstract value $a$ through a
statement $s$, i.e.\ it \emph{deduces} information from $a$ and $s$.
The best abstract transformer is
\[\llbracket s
\rrbracket^\sharp(a)=\bigsqcup \{a'\mid a'\sqsubseteq a, a'\models s\}\] where we write
$a\models s$ if $\forall \vec{y}\in\gamma(a): (s[\vec{y}/\vec{x}]
\equiv \true)$, i.e.\ the constraint $s$ is valid when evaluated over
all values $\vec{y}$ of its variables $\vec{x}$ in the concretisation
of $a$. Any abstract transformer that over-approximates the least
upper bound is a sound transformer and can be used in our algorithm.

In a forward analysis, we can deduce, for example, $\llbracket
x=2(y+z)\rrbracket^\sharp(a)=(-2\leq x\leq 6) \sqcap a$ for $a=(0\leq y \leq 2
\wedge 1 \leq y-z \leq 1)$.
%
We will give more details in Section~\ref{sec:abst}.

>>>>

%-------------------------------------------------------------------------------
\subsection{Abstract Domain}
%-------------------------------------------------------------------------------
SAT solvers operates on the partial assignment domains~\cite{sas12,dhk2013-popl}.  
We present an instantiation of ACDCL as program analyzer that uses relational
domains.  
% 
%In our prototype implementation and for illustrative purposes 
In this paper we use a product domain $\mathcal{B}^b\times\mathcal{R}$ where
$\mathcal{B}$ is the Boolean domain
$\langle\{\true,\false,\bot,\top\},\Rightarrow,\wedge,\vee\rangle$, 
$b$ is the number of boolean variables in the program, 
and $\mathcal{R}$ is a \emph{relational} domain over the 
numerical (bitvector) variables of respective
data types (currently, signed and unsigned integers are supported).
%
We instantiate $\mathcal{R}$ with the template polyhedra domain.

For notational convenience we will denote elements of
$\mathcal{B}^b$ by their concretisations to propositional formulae,
e.g.\ for the boolean variable vector $(g_1,c)$ \rmcmt{what is g1, what is c?} 
with abstract value $(\false,\top)$ we will write $\neg g_1$.

%-------------------------------------------------------------------------------
\paragraph{Template Polyhedra Abstract Domain}
%
The template polyhedra domain~\cite{sriram} represents sets $X$ of
valuations of a vector of variables $\vec{x}$ \rmcmt{not clear what is a
valuation?} such that
$\mat{C}\vec{x}\leq\vec{d}$ for a fixed coefficient matrix $\mat{C}$
and a constant vector $\vec{d}$. The domain of $\vec{d}$ is augmented
by a special element to denote the minimal element $\bot$ of the lattice.
%
The abstraction function is defined as $\alpha(X) = \min \{\vec{d}\mid
\mat{C}\vec{x}\leq\vec{d}, \vec{x}\in X\}$ where $\min$ is
applied component-wise. There are several optimisation-based
techniques for computing the domain
operations~\cite{sriram,GS07b,BJKS15}\rmcmt{like what operations?}. In our implementation we use
the strategy iteration approach available in 2LS~\cite{BJKS15}.
%
The concretisation $\gamma(\vec{d})$ is the set $\{\vec{x}\mid
\mat{C}\vec{x}\leq\vec{d}\}$; $\gamma(\bot)=\emptyset$.
%
For notational convenience we will use conjunctions of linear
inequalities, e.g.\ $x_1\geq 0 \wedge x_1-z\leq 30$ to denote elements
of the \rmcmt{template polyhedra ?} domain. $\true$ is $\top$ and $\false$ is $\bot$.

% \paragraph{Interval Abstract Domain} 
% The {\em Interval} abstract domain is a lattice 
% $ItvDom = (ItvElm, \sle_{I}, \join_{I}, \meet_{I})$, where
% $ItvElm: (Var \rightarrow Itv) \cup \{\bot\}$, and $Itv$ is 
% the set of intervals of type $[l,u]$ over numeric data 
% type with $l \leq u$. The least element is $\bot$ and the 
% greatest element is $\top$ which maps all variables to their
% minimum ($min$) and maximum ($max$) values.  An interval 
% $\langle x, [min, v] \rangle$ is written as $x \leq v$.  The 
% partial order $\sle_{I}$ over elements in the set $Itv$ is 
% given by $I_1 \sle_{I} I_2$ if $I_2$ contains $I_1$.
% A join $(\join_{I})$ of two intervals $\langle x_1 \rightarrow [l_1, u_1], 
% x_1 \rightarrow [l_2, u_2] \rangle$ is an interval 
% $\langle x_1 \rightarrow [min(l_1, l_2), max(u_1, u_2)]$.
% A meet $(\meet_{I})$ of two intervals $\langle x_1 \rightarrow [l_1, u_1], 
% x_1 \rightarrow [l_2, u_2] \rangle$ is an interval 
% $\langle x_1 \rightarrow [max(l_1, l_2), min(u_1, u_2)] \rangle$.
% The galois connection between the concrete environment and intervals is 
% given as follows.
% \[\alpha(x) = \{[inf(x), sup(x)] | x \in Var\} \qquad   \alpha(\emptyset) = \bot \]
% \[\gamma[p,q] = \{c \in \mathbb{Z} | p \leq c \leq q\} \qquad \gamma(\bot) = \emptyset \]
Template polyhedra subsume, for example, interval and octagon domains.
\rmcmt{can we rephrase this statement to say that -- In this paper, we used
interval, octagon and equality ?}

@@

The interval domain is a non-relational domain because a single
inequality only contains a single variable.
%
The octagon domain, however, is relational.  A fundamental difference
between relational and non-relational domains is that relational
domains require the computation of a \emph{closure} in order to obtain
a normal form that is usually required for efficient \rmcmt{or precise?} 
domain operations.  An example of a closure computation for octagonal 
inequalities is shown below.
\[\mathit{closure}((x-y \leq 4) \wedge (y-z \leq 5))=((x-y \leq 4) \wedge (y-z \leq 5) \wedge (x-z \leq 9)) \]  
The closure computes all implied domain constraints.  For octagons,
the closure operation is very expensive (cubic complexity in the
number of program variables).
%
We will see that our model search algorithm performs closure computations lazily,
postponing them to the point where they are actually necessary.\rmcmt{are we
showing this ?}

@@

%-------------------------------------------------------------------------------
\paragraph{Abstract Transformers}
%
%In a typical abstract interpretation based Galois-connection setting
%over an over-approximate domain, every concrete element has a unique
%over-approximate representation in the abstract.  Likewise, every
%concrete transformer is over-approximated by a unique abstract
%transformer.  We now define an abstract deduction transformer.
%
An abstract transformer transforms an abstract value $a$ through a
statement $s$, i.e.\ it \emph{deduces} information from $a$ and $s$.
The best abstract transformer is \rmcmt{is this equation updated ?}
\[\llbracket s
\rrbracket^\sharp(a)=\bigsqcup \{a'\mid a'\sqsubseteq a, a'\models s\}\] where we write
$a\models s$ if $\forall \vec{y}\in\gamma(a): (s[\vec{y}/\vec{x}]
\equiv \true)$, i.e.\ the constraint $s$ is valid when evaluated over
all values $\vec{y}$ of its variables $\vec{x}$ in the concretisation
of $a$. Any abstract transformer that over-approximates the least
upper bound is a sound transformer and can be used in our algorithm.

In a forward analysis, we can deduce, for example, $\llbracket
x=2(y+z)\rrbracket^\sharp(a)=(-2\leq x\leq 6) \sqcap a$ for $a=(0\leq y \leq 2
\wedge 1 \leq y-z \leq 1)$.
%
A detailed description of abstract transformer is presented in Section~\ref{sec:abst}.
